The host galaxy of the short GRB111117A at z = 2.211. \thanks{Based on observations collected at the European Southern Observatory, Paranal, Chile, Program ID: 088.A-0051 and 091.D-0904.}

### 2017-05-24 GRB111117A.tex
Test tex update
Confirm progenitor seperation and push calc code

\abstract{
Short $\gamma$-ray bursts (sGRBs) and their hosts are notoriously difficult to localize. We here present evidence of the, to date, highest redshift high-confidence short GRB, located at $z=2.211$. This value supersedes a previously estimated, lower redshift value. Spectroscopic observations along with a wealth of imaging data allow us to place tight constraints on the nature of the host galaxy at an unprecedented distance. 
The rest frame X-ray derived hydrogen column density is high compared to a complete sample of sGRBs and seems to follow the evolution with redshift as traced by the hosts of long GRBs (lGRBs). This could indicate that the part of the sGRB population hosted in late-type galaxies, live in similar environments to that of lGRB hosts. 
The host is in the brighter end of the expected host brightness distribution at $z = 2.211$, but only 30 per cent of the distribution should be missed due to host faintness at this redshift. This indicates that we are not missing redshifts for a dominant fraction of the sGRB hosts at this redshift. The redshift of GRB111117A is evidence against a lognormal delay-time model for sGRBs through the predicted redshift distribution of sGRBs, which is very sensitive to high-z sGRBs.
From the age of the universe at explosion time, an initial progenitor separation of a$_0 < 3.1 $R$_\odot$ is required for the case of a binary neutron star (NS) system. This puts constraints on the progenitor system evolution up to the time of explosion.
}

@@

At $z = 2.211$, the age of the universe is 2.97 Gyr. If the progenitor systems of sGRBs are the merger of two neutron stars, this sets a hard upper limit to the coalescence timescale for such a system. In the absence of other mechanisms, the timescale of the orbital decay of the system is set by the energy loss due to gravitational waves, which in turn is set by the mass of constituent compact objects and the separation of the two \citep{Postnov2014}. If we assume that the formation timescale of the first galaxies is short compared to the time since the big bang \citep{Richard2011} and if we assume a mass of 1.4 M$_\odot$ for each of the neutron stars at the time of system formation, this places a limit on the initial separation of the two neutronstars of a$_0 < 3.1 $R$_\odot$, where a$_0$ is the initial separation. Because we have the age of the stellar population, by using this as a limit to the orbital decay time we can put tighter constraints, a$_0 < 1.7 $R$_\odot$, but this does not account for the possibility there could be an underlying stellar population of older stars from a previous star-formation epoch, which is why we consider a$_0 < 3.1 $R$_\odot$ as a hard upper limit. The delay time between formation and explosion is well accomodated by the models of \citet{Belczynski2006}, although the longest formation channels are excluded. This is especially true given the late type nature of the host \citep{OShaughnessy2008}.

@@

Using the age of the universe at the time of explosion allows us to set constraints on the maximal separation between the engine constituents at the time of formation. We find that the maximal separation for two neutron stars at formation time is a$_0 < 3.1 $R$_\odot$, which excludes some of the formation channels with the longest timescales.

>>>>

\abstract{
Short $\gamma$-ray bursts (sGRBs) and their hosts are notoriously difficult to localize. We here present evidence of the, to date, highest redshift high-confidence short GRB, located at $z=2.211$. This value supersedes a previously estimated, lower redshift value. Spectroscopic observations along with a wealth of imaging data allow us to place tight constraints on the nature of the host galaxy at an unprecedented distance. 
The rest frame X-ray derived hydrogen column density is high compared to a complete sample of sGRBs and seems to follow the evolution with redshift as traced by the hosts of long GRBs (lGRBs). This could indicate that the part of the sGRB population hosted in late-type galaxies, live in similar environments to that of lGRB hosts. 
The host is in the brighter end of the expected host brightness distribution at $z = 2.211$, but only 30 per cent of the distribution should be missed due to host faintness at this redshift. This indicates that we are not missing redshifts for a dominant fraction of the sGRB hosts at this redshift. The redshift of GRB111117A is evidence against a lognormal delay-time model for sGRBs through the predicted redshift distribution of sGRBs, which is very sensitive to high-z sGRBs.
From the age of the universe at explosion time, an initial progenitor separation of a$_0 < 3.1 $R$_\odot$ is required for the case of a binary neutron star (NS) system. This puts constraints on the progenitor system evolution up to the time of explosion.

}

@@

At $z = 2.211$, the age of the universe is 2.97 Gyr. If the progenitor systems of sGRBs are the merger of two neutron stars, this sets a hard upper limit to the coalescence timescale for such a system. In the absence of other mechanisms, the timescale of the orbital decay of the system is set by the energy loss due to gravitational waves, which in turn is set by the mass of constituent compact objects and the separation of the two \citep{Postnov2014}. If we assume that the formation timescale of the first galaxies is short compared to the time since the big bang \citep{Richard2011} and if we assume a mass of 1.4 M$_\odot$ for each of the neutron stars at the time of system formation, this places a limit on the initial separation of the two neutronstars of a$_0 < 3.2 $R$_\odot$, where a$_0$ is the initial separation. Because we have the age of the stellar population, by using this as a limit to the orbital decay time we can put tighter constraints, a$_0 < 2.1 $R$_\odot$, but this does not account for the possibility there could be an underlying stellar population of older stars from a previous star-formation epoch, which is why we consider a$_0 < 3.2 $R$_\odot$ as a hard upper limit. The delay time between formation and explosion is well accomodated by the models of \citet{Belczynski2006}, although the longest formation channels are excluded. This is especially true given the late type nature of the host \citep{OShaughnessy2008}.

@@

Using the age of the universe at the time of explosion allows us to set constraints on the maximal separation between the engine constituents at the time of formation. We find that the maximal separation for two neutron stars at formation time is a$_0 < 3.2 $R$_\odot$, which excludes some of the formation channels with the longest timescales.

We make all the data, code and calculation related to the paper along with the paper itself available at \url{https://github.com/jselsing/GRB111117A}.
